High-order dynamic neural correlations reflect naturalistic processing in humans

### 2019-05-08 paper/main.tex
almost done with the draft... small tweaks still needed

\subsubsection*{Inter-subject dynamic correlations}
Equation~\ref{eqn:timecorr} provides a means of taking a single
observation matrix, $\mathbf{X}$ and estimating the dynamic
correlations from moment to moment, $\mathbf{Y}$.  Suppose that one
has access to a set of multiple observation matrices that reflect the
same phenomenon.  For example, one might collect neuroimaging data
from several experimental participants, as each participant performs
the same task (or sequence of tasks).  Let $\mathbf{X}_1$,
$\mathbf{X}_2$, ..., $\mathbf{X}_P$ reflect the $T$ by $F$ observation
matrices for each of $P$ participants in an experiment.  We can use
\textit{inter-subject functional connectivity}~\citep[ISFC;
][]{SimoEtal16} to compute the degree of stimulus-driven correlations
reflected in the multi-participant dataset at a given timepoint $t$
using:
\begin{align}
\bar{\mathbf{C}}(t) = M\left(R\left(\frac{1}{2P} \sum_{i=1}^P
  Z\left(Y_i(t)\right)^\mathrm{T} + Z\left(Y_i,(t)\right)\right)\right),
\end{align}
where $M$ extracts and vectorizes the diagonal and upper triangle of a symmetric
matrix, $Z$ is the Fisher $z$-transformation~\citep{Zar10}:
\begin{align}
Z(r) = \frac{\log(1+r) - \log(1-r)}{2}
\end{align}
$R$ is the inverse of $Z$:
\begin{align}
R(z) = \frac{\exp(2z - 1)}{\exp(2z + 1)},
\end{align}
and $\mathbf{Y}_i(t)$ denotes the correlation matrix
(Eqn.~\ref{eqn:corr}) between each column of $\mathbf{X}_i$ and each
column of the average observations from all \textit{other}
participants, $\bar{\mathbf{X}}_{ \backslash i}$:
\begin{align}
  \bar{\mathbf{X}}_{ \backslash i} = R\left(\frac{1}{P-1}\sum_{i \in
  \backslash i} Z\left( \mathbf{X}_i \right) \right),
\end{align}
where $ \backslash i$ denotes the set of all participants other than
participant $\i$. In this way, the $T$ by $\left( \frac{F^2 - F}{2}
  + F \right)$
matrix $\bar{\mathbf{C}}$ is the time-varying extension of the ISFC
approach developed by \cite{SimoEtal16}.

@@

We applied timecorr, using delta and gaussian kernels
Fig.~\ref{fig:weights}) to each of these 
synthetic datasets, then correlate each recovered
correlation matrix with the ground truth.  We explore how recovery varies
with the kernel and the specific structure of the data. For the
ramping synthetic dataset (Fig.~\ref{fig:synthetic_data},  C.)  and for the
block synthetic dataset (Fig.~\ref{fig:synthetic_data},  D.)  we made further
comparisons of the timecorr recovered correlation matrices. We
compared the ramping recovered correlation matrices to only the first random covariance matrix $K_{1}$
(First, Fig.~\ref{fig:synthetic_data},  C.) and to only the last
random covariance matrix $K_{2}$ (Last, Fig.~\ref{fig:synthetic_data},
C.) from Equation~\ref{eqn:ramping}. We also compared the block recovered correlation matrices in to
the block specific covariance matrix (Block 1-5,
Fig.~\ref{fig:synthetic_data},  D.).

@@

\begin{itemize}

\item Decoding accuracy best for level 2 data for all but rest
  condition.  Could be that the brain is 2nd order or that fMRI can
  only reliable give 2nd order.  
\item multiple timescale representations (a la Hasson group) implies
first-order network interactions.  higher-order interactions imply
generalizations between interacting representations (e.g. mirrored
schema, a la Norman/Baldassano/Hasson).  possibly cite NTB 2013
science review, using as evidence that this is where the field is
going (voxels $\rightarrow$ patterns (L0) $\rightarrow$ interactions (L1) $\rightarrow$ higher
order patterns (L2$+$).

@@

\item related approaches: sliding window, phase-based correlations,
within-ROI spatial correlations at each timepoint, granger
causality, other explit models (e.g. virtual brain).
\item other applications: molecular interactions (protein folding?),
diagnosis (e.g. psychiatric disorders as network flow problems--
gratton work?), social network dynamics (e.g. financial markets,
social media interactions)
\end{itemize}

Dynamic interactions between brain structures underlie our thoughts.
We developed a model of neural dynamics that takes in a feature
timeseries and outputs estimated first-order network dynamics (i.e.,
dynamic functional correlations), second-order network dynamics
(reflecting homologous networks that dynamically form and disperse),
and higher-order network dynamics (up to tenth-order dynamic
correlations).  After validating our model using synthetic data, we
applied the approach to an fMRI dataset
~\citep{SimoEtal16} in which participants listened to an audio recording of a story, as well as scrambled versions of the same story (where the scrambling was applied at different temporal scales).  We trained classifiers to take the output of the model and decode the timepoint in the story (or scrambled story) that the participants were listening to. We found that, during each of the listening conditions in the
experiment, classifiers that incorporated up to second-order correlations
yielded consistently higher accuracy than classifiers trained only on
lower-order patterns (Fig.~\ref{fig:decoding_level},  A.).  (Incorporating higher-order correlations did
not further improve decoding accuracy.)  By contrast, these
higher-order correlations were not necessary to support (minimally
above chance) decoding during a control rest condition.  This suggests
that the cognitive processing that supported the listening conditions
involved second-order (or higher) network dynamics.

Although we found decoding accuracy was best for second-order network dynamics for all but rest
  condition, it is unclear if this is a product of the brain or the
  data collection technique .  It could be that the brain is
  second-order or it could be that fMRI can
  only reliably give second-order interactions.  Exploring this method
  with other data collection technique will be important to
  disentangle this question.

\subsection*{Concluding remarks}
the universe is complicated and we need scalable approaches to
studying how the pieces are interacting to make sense of it.  one
small step for mankind, and so on.

>>>>

\subsubsection*{Inter-subject dynamic correlations}
Equation~\ref{eqn:timecorr} provides a means of taking a single
observation matrix, $\mathbf{X}$ and estimating the dynamic
correlations from moment to moment, $\mathbf{Y}$.  Suppose that one
has access to a set of multiple observation matrices that reflect the
same phenomenon.  For example, one might collect neuroimaging data
from several experimental participants, as each participant performs
the same task (or sequence of tasks).  Let $\mathbf{X}_1$,
$\mathbf{X}_2$, ..., $\mathbf{X}_P$ reflect the $T$ by $F$ observation
matrices for each of $P$ participants in an experiment.  We can use
\textit{inter-subject functional connectivity}~\citep[ISFC;
][]{SimoEtal16} to compute the degree of stimulus-driven correlations
reflected in the multi-participant dataset at a given timepoint $t$
using:
\begin{align}
\bar{\mathbf{C}}(t) = M\left(R\left(\frac{1}{2P} \sum_{i=1}^P
  Z\left(Y_i(t)\right)^\mathrm{T} + Z\left(Y_i,(t)\right)\right)\right),
\end{align}
where $M$ extracts and vectorizes the diagonal and upper triangle of a symmetric
matrix, $Z$ is the Fisher $z$-transformation~\citep{Zar10}:
\begin{align}
Z(r) = \frac{\log(1+r) - \log(1-r)}{2}
\end{align}
$R$ is the inverse of $Z$:
\begin{align}
R(z) = \frac{\exp(2z - 1)}{\exp(2z + 1)},
\end{align}
and $\mathbf{Y}_i(t)$ denotes the correlation matrix
(Eqn.~\ref{eqn:corr}) between each column of $\mathbf{X}_i$ and each
column of the average observations from all \textit{other}
participants, $\bar{\mathbf{X}}_{ \backslash i}$:
\begin{align}
  \bar{\mathbf{X}}_{ \backslash i} = R\left(\frac{1}{P-1}\sum_{i \in
  \backslash i} Z\left( \mathbf{X}_i \right) \right),
\end{align}
where $ \backslash i$ denotes the set of all participants other than
participant $i$. In this way, the $T$ by $\left( \frac{F^2 - F}{2}
  + F \right)$
matrix $\bar{\mathbf{C}}$ is the time-varying extension of the ISFC
approach developed by \cite{SimoEtal16}.

@@

We applied timecorr, using delta and gaussian kernels
Fig.~\ref{fig:weights}) to each of these 
synthetic datasets, then correlated each recovered
correlation matrix with the ground truth.  We repeated this process 10
times and explored how recovery varies
with the kernel and the specific structure of the data. For the
ramping synthetic dataset (Fig.~\ref{fig:synthetic_data},  C.)  and for the
block synthetic dataset (Fig.~\ref{fig:synthetic_data},  D.)  we made further
comparisons of the timecorr recovered correlation matrices. We
compared the ramping recovered correlation matrices to only the first random covariance matrix $K_{1}$
(First, Fig.~\ref{fig:synthetic_data},  C.) and to only the last
random covariance matrix $K_{2}$ (Last, Fig.~\ref{fig:synthetic_data},
C.) from Equation~\ref{eqn:ramping}. We also compared the block recovered correlation matrices in to
the block specific covariance matrix (Block 1-5,
Fig.~\ref{fig:synthetic_data},  D.).

@@

@@

Although dynamic interactions between brain structures underlie our
thoughts, the ways in which we can study these patterns are
limited. One challenge in studying dynamic interactions is the
computational resources required to calculate higher-order correlations. 
We developed a computationally tractable model of network dynamics (Fig.~\ref{fig:methods_fig}) that takes in a feature
timeseries and outputs approximated first-order dynamics (i.e.,
dynamic functional correlations), second-order dynamics
(reflecting homologous networks that dynamically form and disperse),
and higher-order network dynamics (up to tenth-order dynamic
correlations).

We first validated our model using synthetic data, and explored how
recovery varied with different underlying data structures  and kernels.   We then 
applied the approach to an fMRI dataset
~\citep{SimoEtal16} in which participants listened to an audio recording of a story, as well as scrambled versions of the same story (where the scrambling was applied at different temporal scales).  We trained classifiers to take the output of the model and decode the timepoint in the story (or scrambled story) that the participants were listening to. We found that, during each of the listening conditions in the
experiment, classifiers that incorporated up to second-order correlations
yielded consistently higher accuracy than classifiers trained only on
lower-order patterns (Fig.~\ref{fig:decoding_level},  A.), and
incorporating more higher-order correlations did
not further improve decoding accuracy.  By contrast, these
higher-order correlations were not necessary to support (minimally
above chance) decoding during a control rest condition.  This suggests
that the cognitive processing that supported the listening conditions
involved second-order (or higher) network dynamics.

Although we found decoding accuracy was best when incorporating
higher-order network dynamics for all but rest
  condition, it is unclear if this is a product of the brain or the
  data collection technique .  It could be that the brain is
  second-order or it could be that fMRI can
  only reliably give second-order interactions. Exploring this method
  with other data collection technique will be important to
  disentangle this question.

  \subsection*{Concluding remarks}

How can we better understand how these brain patterns changing over
time? How to quantify the potential network dynamics that we might be
driving these changes? One way to judge the techniques of the future is
to look at the trajectory of the fMRI field so far has taken so far
(Fig.~\ref{fig:methods_fig}).  The field started with 
univariate activation, measuring the average activity for each voxel.
Analyses of multivariate activation followed, looking at spatial patterns of
activity over voxels. Next, correlations of activity were explored, first
with measures like resting connectivity that take temporal correlation
between a seed voxel and all other voxels then with full connectivty
that measure all pairwise correlations.  Additionally, this path of increasing
complexity also moved from static to dynamic measurements.  One
logical next step in this trajectory would be dynamic higher-order
correlations. We have created a method 
to support these calculations.
